Praise the people or praise the place?

### 2025-04-15 03-praise-the-people-or-praise-the-place.qmd
update Kappa'

---
title: "Praise the people or praise the place?"
subtitle: Upper tail human capital in electrifying Sweden
code-repo: "Access the code, data, and analysis at <https://github.com/j-jayes/who-is-who-etl>"
author:
  - name: Jonathan Jayes
    email: jonathan.jayes@ekh.lu.se
    orcid: 0000-0003-4967-4869
    affiliations: 
      - id: lu
        name: Lund University Economic History Department
abstract: |
  This paper explores the transformative role of Artificial Intelligence in economic history research, through the examination of the skilled workforce in Sweden's electricity sector during the mid-20th century. By ingesting and analyzing two unique data sources—a set of biographical dictionaries and an industrial catalogue—the study investigates the origins and trajectories of high-skilled workers in electricity-related occupations in Sweden. Contrary to previous assumptions that local talent predominantly drove the technological advancements, the findings reveal a bifurcated labor market. Highly educated and skilled professionals, crucial in overseeing and advancing the electricity sector, frequently migrated for education and career opportunities, often traveling significant distances from their birthplaces. In contrast, local talent primarily occupied the burgeoning middle-skilled roles within the sector.

  New abstract.

  This paper investigates the origins and career trajectories of high-skilled workers in electricity-related occupations in Sweden during the mid-20th century. By ingesting and analyzing two unique data sources – a set of biographical dictionaries and an industrial catalogue – I ask where these workers came from, what and where they studied, and how international experience impacted their career paths.

  Jayes, Enflo and Molinder (2024) find that medium-skilled electricity-related jobs were filled by workers who were born near to these jobs. However, descriptive statistics about these high-skilled workers reveal a bifurcated labor market. Highly educated and skilled professionals, crucial in overseeing and advancing the electricity sector, frequently migrated for education and career opportunities, often traveling significant distances from their birthplaces to a small set of higher education institutions in Sweden and abroad. After their studies, Swedish engineers moved to opportunity, ending up on average ten times further from their birthplace than the median worker in 1930. Tentative findings on international experience show that engineers worked abroad at higher rates than comparable occupations, far more than doctors or dentists.

date: 12 January, 2024
thanks: |
  Thank you to seminar participants at the Copenhagen Business School Department of Strategy and Innovation PhD Seminar  and the HEDG Group at the University of Southern Denmark for valuable feedback on this work.
format:
  hikmah-pdf: default
  hikmah-manuscript-pdf: default
# format:
#   docx:
#     toc: true
#     number-sections: true
#     highlight-style: github
execute: 
  echo: false
  warning: false
  message: false
bibliography: bibliography.bib
---

@@

Economic history as a discipline is set to benefit in at least three ways from the nascent AI revolution. Novel sources of data become available as new tools make them readable to machines, analysis in new ways is possible with new kinds of algorithms, and the time-cost of asking certain kinds of questions decreases, opening up new avenues for research. In this paper, I make an attempt to leverage these benefits to answer the question, 'who were the high skilled workers in electricity related occupations in Sweden, and where did they come from?' I detail the process of structuring and analyzing two novel data sources, a set of biographical dictionaries and an industrial catalogue, in order to answer this question.

In the quest to understand the dynamics of economic development and technological advancement, previous research by this author and his supervisors shed light on the transformative impact of early electricity access in Sweden. "Power for progress: The impact of electricity on individual labor market outcomes" [@jayes_power_2024] revealed how the advent of electricity in certain parishes led to positive economic outcomes: a boost in income levels, reduced inequality, and the maintenance of employment levels despite the advent of labor-saving technology. A particularly striking observation was the tendency of workers in these early electrified parishes to remain in their birthplaces, hinting at a newfound economic vibrancy stemming from the income spillovers into sectors not affected by the new technology.

Building on these insights, the present paper delves deeper into the human aspect of this technological revolution. It poses a critical question: Who were the key figures driving this change? Were they local talents nurtured by the opportunities at hand, or did they represent a wave of skilled individuals drawn from afar, lured by the pioneering spirit of these early electrified areas?

To answer this, the investigation leverages two novel and rich data sources. The first, *Vem är Vem*, is a comprehensive set of biographical dictionaries containing the profiles of 75,000 notable Swedes active between 1945 and 1968. The second, the "Svensk Industrikalender" or Swedish Industrial Calendar of 1947, offers an exhaustive catalogue of industrial firms, detailing their activities, workforce, and financial metrics.

I digitize and structure these sources in order to analyze the changing patterns of the Swedish labour market in the middle of the 20th century in light of electrification. The findings challenge my prior expectations. Contrary to the belief that local talent pools predominantly fueled the technological boom, I observe a pattern of geographical mobility among the highly educated and skilled professionals in electricity-related fields. These individuals, pivotal to overseeing and advancing the electricity sector, often sought education and opportunities far from their origins. This suggests a bifurcated labor market: local talent predominantly filled the burgeoning middle-skilled roles within the electricity sector, while the top-tier skilled professionals were more transient, moving towards educational and occupational opportunities. This paper explores the implications of this labor market structure for the economic development patterns witnessed during Sweden's second Industrial Revolution.

@@

The paper is laid out as follows: the current research question is placed in context, the sources are explained, followed by their digitization and structuring process. I then discuss the classification task involved in assigning engineers to a sector based on their occupational trajectories. I then lay out some descriptive statistics and tentative findings regarding the patterns of movement for the high skilled electricity related workers, compared with other professionals I observe in the biographical dictionaries.

## Related Literature

@@

@@

Some interesting papers that incorporate prediction and classification problems include @Bandiera2020ceo, who classify CEO behavior by collecting high-frequency diary information and then use a machine learning algorithm to classify CEOs into 'leaders' and 'managers' by the content of their meetings. @koschnick2023breaking uses a machine learning topic model to classify each paper by the universe of all students at English universities in the seventeenth and early eighteenth century to calculate a measure of how innovative the paper was; how it differed from the papers before it in the field and how similar the papers afterwards were. @DahlVedel2024 in a paper titled "Breaking the HISCO Barrier: AI and Occupational Data Standardization" apply a neural network to the task of classifying an occupational description and benchmark their results against human labelling to show that the neural network achieved comparable accuracy with human labelling and involved an order of magnitude fewer human hours.

@@

## Source material

@@

@@

In order to analyze both the biographical dictionaries and industrial catalogue, we need to bend the text into a machine readable structure. This process is not complicated, but somewhat involved. It includes breaking each component of the source up (e.g. each biography or company record), extracting the pertinent information from each record, storing each value with its associated key, and then saving this information in a way that is easy to analyze and aggregate.

@@

I make use of the computational backend of Chat-GPT, a model called GPT-3.5-turbo from OpenAI to structure the information from the dictionaries and catalogue into a JSON format that I can analyze, step 3, as shown in @fig-data-collection-process. Many other LLMs, some open source, are available. I have chosen GPT-3.5-turbo as it is simple to interact with it in the programming language Python, and because doing so is relatively affordable for contained workloads such as this, when compared to hosting such a model on your own, beefy, costly computer.

@@

The GPT-3.5-turbo model which I make use of is a LLM which has been trained on all of Wikipedia and Wikidata, among other training material. These two sources contain the same information, but in a different format, as shown in the adapted extracts below; the text and @fig-wiki mimic the kind of material that the model I use is trained on.

As the base model underlying GPT-3.5-turbo is pre-trained to predict the next token on this kind of data, it has developed the ability through repeated exposure to this kind of biographic data to produce structured information from free text, and likewise construct natural sounding text from pieces of structured information. With the addition of *JSON mode* to the model, it is able to output structured information in a reliable manner as detailed in @openai2024text.

@@

Due to the large amount of material that GPT-3.5-turbo has been exposed to in training, it is familiar with the kinds of biographical text that I want it to structure. In order to draw on its familiarity with this kind of material, I need to provide it context about the biographical text it is being passed, and ensure that the model returns output in a useful way. I explain these prompts below.

@@

Next I explain the kinds of information that I want it to extract, and the exact format that I want it in. I do this by specifying a JSON schema, with fields that it must return and specifications for the kind of data in each field. An except of the schema is shown in @fig-schema.

```{r}
#| label: fig-schema
#| fig-cap: "Excerpt of structuring schema"
#| fig-width: 3
#| out-width: 3in

knitr::include_graphics("assets/schema-2.png")
```

@@

`structure_prompt = "Task: read the schema and return RFC compliant JSON information about the Swedish individuals from the 1950 biographical dictionary 'Vem är Vem' that is provided below. Use a numeric index for each biography in your JSON output and return information about all of them, including all career information available. Keep the biographic descriptions in Swedish and remove any abbreviations based on your knowledge, e.g. 'fil. kand.' is 'filosofie kandidat', and 'Skarab. l.' is 'Skaraborgs Län'. Put years in full based on context. Put dates in dd/mm/yyyy format where possible. If there is no information for a key, leave it out. If there is no information for a required key, put NULL as the value.`

@@

person_info <- tribble(
  ~key, ~value,
  "full_name", "Nordell, Axel Verner",
  "location", "Motala, Östergötland",
  "occupation", "Civilingenjör, kraftverksdirektör",
  "birth_date", "15/08/1881",
  "birth_place", "S. Möckleby, Kalmar",
  "birth_parents", "Gustaf N. and Amanda Seillergren",
  "birth_latitude", "56.35646300000001",
  "birth_longitude", "16.420155",
  "education_degree", "Studentexamen",
  "education_year", "1899",
  "education_institution", "Lunds högre allmänna läroverk",
  "education_latitude", "55.7046601",
  "education_longitude", "13.1910073"
)

@@

## Clustering of firms and classification of occupations

The next task required to understand where these individuals came from and what kinds of firms they were drawn to involved grouping the firms from the catalogue and then classifying each individual in the biographical dictionaries into a particular occupation.

For both of these tasks, I leaned on the tools of text embeddings, and a combination of unsupervised machine learning and advanced language processing techniques. I made use of a text embedding model trained on Swedish language text by the @bert-base-swedish-cased. It is an adaptation of the breakthrough BERT model, introduced by Google Research in 2018 [@google2016bert]. The advantage of this KB Lab model is that it has been trained on a selection of Swedish data, including books, news reports, and internet forums. Hence it is able to score the similarity of Swedish business descriptions and occupational titles.

@@

In @jayes_power_2024, we grouped occuaptions into three categories; direct electricity jobs (e.g. electricians), indirect electricity jobs that could benefit from electric motors (e.g. textile workers), and all other jobs. We made this classification based on the occupational title listed in the 1930 census alone. These titles are frequently used in economic history, along with a schema that classifies each title according to a list of possible titles and occupational descriptions. A widely used example is the Historical International Standard Classification of Occupations, or HISCO, defined originally by @vanLeeuwen2002HISCO. A wealth of mappings have been created that link an occupational string like *civilingenjör* to HISCO code 022, civil engineers, described as:

@@

While this classification process used to involve a large amount of human hours, several new methods that use machine learning have emerged to make this process more efficient. @DahlVedel2024 use a neural network to classify occupational strings using a supervised machine learning process based on the occupational titles in several waves of Danish census data. @merouani2023innovation uses an unsupervised approach based on text embeddings to classify french occupational titles according to the HISCO schema, using a clever compartmentalization strategy to improve classification when individuals hold multiple titles.

In this paper I take a slightly different approach, making use of the additional data from the occupational trajectories of each individual; where they worked and what position they held at each firm or institution. This additional data helps me go beyond just a single occupational string and classify the sector that each engineer worked in. In many instances, the description of the individual in the biographical data relates to their education, rather than what they have done with it.

@@

### Clustering firms by business type and geography

For grouping firms by business type, I again used the k-means clustering algorithm, to categorize firms into 20 distinct categories. This algorithm works by partitioning the data into k distinct clusters based on features (in this case, text embeddings derived from company descriptions) [@jurafsky2023speech]. I experimented and decided on 20 clusters and manually labelled each cluster based on the firm names and business operations of the firms in the different clusters.

>>>>

---
title: "Praise the people or praise the place?"
subtitle: Upper tail human capital in electrifying Sweden
code-repo: "Access the code, data, and analysis at <https://github.com/j-jayes/who-is-who-etl>"
author:
  - name: Jonathan Jayes
    email: jonathan.jayes@ekh.lu.se
    orcid: 0000-0003-4967-4869
    affiliations: 
      - id: lu
        name: Lund University Economic History Department
abstract: |  
  This paper investigates the origins and career trajectories of high-skilled workers in electricity-related occupations in Sweden during the mid-20th century. By ingesting and analyzing two unique data sources – a set of biographical dictionaries and an industrial catalogue – I ask where these workers came from, what and where they studied, and how international experience impacted their career paths.

  Jayes, Enflo and Molinder (2024) find that medium-skilled electricity-related jobs were filled by workers who were born near to these jobs. However, descriptive statistics about these high-skilled workers reveal a bifurcated labor market. Highly educated and skilled professionals, crucial in overseeing and advancing the electricity sector, frequently migrated for education and career opportunities, often traveling significant distances from their birthplaces to a small set of higher education institutions in Sweden and abroad. After their studies, Swedish engineers moved to opportunity, ending up on average ten times further from their birthplace than the median worker in 1930. In terms of occupational mobility, we do not see a similar pattern to the middle-skill electricity related jobs among engineers; there is no clear pattern for engineers to have lower status parents than other occupations in the biographical dictionaries.

date: today
thanks: |
  Thank you to seminar participants at the Copenhagen Business School Department of Strategy and Innovation PhD Seminar  and the HEDG Group at the University of Southern Denmark for valuable feedback on this work.
format:
  hikmah-pdf: default
  hikmah-manuscript-pdf: default
# format:
#   docx:
#     toc: true
#     number-sections: true
#     highlight-style: github
execute: 
  echo: false
  warning: false
  message: false
bibliography: bib.bib
---

@@

Economic history as a discipline is set to benefit in at least three ways from the nascent AI revolution. Novel sources of data become available as new tools make them readable to machines, analysis in new ways is possible with new kinds of algorithms, and the time-cost of asking certain kinds of questions decreases, opening up new avenues for research. In this paper, I make an attempt to leverage these benefits to answer the question, 'who were the high skilled workers in electricity related occupations in Sweden, and where did they come from?' I detail the process of structuring and analyzing a novel data source; a set of biographical dictionaries in order to answer this question.

In the quest to understand the dynamics of economic development and technological advancement, previous research by this author and his supervisors shed light on the transformative impact of early electricity access in Sweden. "Power for progress: The impact of electricity on individual labor market outcomes" [@jayes_power_2024] revealed how the advent of electricity in certain parishes led to positive economic outcomes: a boost in income levels, reduced inequality, and the maintenance of employment levels despite the advent of labor-saving technology. There were two findings that stood and which I seek to explore further here. First was the tendency of workers in these early electrified parishes to remain in their birthplaces, hinting at a newfound economic dynamism stemming from the income spillovers into sectors not affected by the new technology. Second, we found that electrification allowed those without a great deal of formal education to take on new jobs that emerged in the wake of the new technology - like electricians and electrical linemen - leading to a more equal distribution of income in these areas.

Building on these insights, the present paper looks deeper into the human aspect of this technological revolution. It poses a question: Who were the key figures driving this change? Were they local talents, or did they represent a wave of skilled individuals drawn from afar, drawn in by the dynamism of these early electrified areas? We can also ask about their occupational mobility, and whether they were less likely to have parents in high status occupations than other individuals in the biographical dictionaries.

To answer this, the investigation leverages a novel data source. *Vem är Vem*, is a comprehensive set of biographical dictionaries containing the profiles of 75,000 notable Swedes active between 1945 and 1968.

I digitize and structure this source in order to analyze the changing patterns of the Swedish labour market in the middle of the 20th century in light of electrification. The findings challenge my prior expectations. Contrary to the belief that local talent pools predominantly fueled the technological boom, I observe a pattern of geographical mobility among the highly educated and skilled professionals in electricity-related fields. These individuals, pivotal to overseeing and advancing the electricity sector, often sought education and opportunities far from their origins. This suggests a bifurcated labor market: local talent predominantly filled the burgeoning middle-skilled roles within the electricity sector, while the top-tier skilled professionals were more transient, moving towards educational and occupational opportunities. This paper explores the implications of this labor market structure for the economic development patterns witnessed during Sweden's second Industrial Revolution.

@@

The paper is laid out as follows: the current research question is placed in context, the source is explained, followed by the digitization and structuring process. I then discuss the classification task involved in assigning engineers to a sector based on their occupational trajectories. I then lay out some descriptive statistics and tentative findings regarding the patterns of movement for the high skilled electricity related workers, compared with other professionals I observe in the biographical dictionaries.

## Context

### Introduction

The electrification of Sweden stands as a pivotal chapter in its economic history, transforming industry, boosting productivity, and reshaping labor markets [cite: 1 (Schon (2000) Electricity, technological change and productivity in Swedish industry, 1890-1990.pdf)]. Lennart Schön (2000) highlights electricity's central role, particularly given Sweden's abundant hydropower resources and lack of fossil fuels, noting its rapid adoption spurred industrial development from the 1890s onwards [cite: 1 (Schon (2000) Electricity, technological change and productivity in Swedish industry, 1890-1990.pdf)]. Recent research, leveraging the quasi-experimental rollout of the electrical grid connecting Sweden's early major hydropower plants (the "Western Line"), has provided granular insights into its impact. Jayes, Molinder, and Enflo (2024) demonstrate that parishes gaining early access to electricity saw significant individual income gains, particularly for lower-income workers and those with only primary education, leading to reduced local income inequality [cite: 1 (02-power-for-progress.pdf)]. This suggests electricity acted as a labor-supporting technology, improving outcomes for incumbents, many of whom remained in their birth parishes despite the technological shift [cite: 1 (02-power-for-progress.pdf)].

However, this picture primarily reflects the experience of the broader, often lower-skilled, labor force. Preliminary work using biographical dictionaries of notable Swedes ("Praise the People," Jayes et al. 2024) suggests a contrasting pattern for high-skilled workers, particularly engineers [cite: 1 (03-praise-the-people-or-praise-the-place.pdf)]. These individuals, crucial for designing, implementing, and managing the new electrical systems, appear to have been geographically highly mobile, often pursuing education and career opportunities far from their places of birth [cite: 1 (03-praise-the-people-or-praise-the-place.pdf)]. This raises a critical question: If early electrification created distinct local advantages, did it attract a systematically *different* type of high-skilled engineer compared to other, albeit slightly later-electrifying, areas? Were the engineers moving *to* the pioneering "Western Line" parishes different in their backgrounds from those moving *to* otherwise comparable parishes within the broader region influenced by electrification?

This paper investigates this question by comparing the background characteristics of engineers who migrated into the early-electrifying "Western Line" parishes with those who migrated into control parishes within the 300km band identified by Jayes, Molinder, and Enflo (2024) [cite: 1 (02-power-for-progress.pdf)]. Using biographical data detailing origins, education (institution and field), parental occupation (via HISCO classification), age at move, family formation, and international experience, we examine whether distinct observable traits characterized engineers attracted to the earliest electrified locations.

Our analysis reveals a striking degree of similarity between the two groups of engineers. Across dimensions such as educational institution, field of study, and social background (proxied by paternal occupation), we find no significant differences between engineers moving to Western Line parishes and those moving to control parishes. This suggests that, at the high-skill level, engineers were a highly mobile group attracted broadly to the economically dynamic region undergoing electrification, rather than being selectively sorted based on these characteristics to the very first parishes gaining grid access. The primary discernible difference is a tendency for engineers moving to the Western Line parishes to have slightly more prevalent prior work experience in the United States. No similar distinction was found for experience gained in Europe.

This finding contributes to our understanding of high-skilled labor markets during periods of major technological change. It implies that while electrification generated localized benefits for the general workforce, the specific *timing* of access within a region undergoing transformation did not strongly select for engineers with particular educational or social backgrounds. Instead, the broader dynamism of the region attracted a mobile pool of talent, distinguished perhaps only subtly by the specific *type* of international experience relevant to the emerging technological frontier, in this case, American know-how.

### Literature Review

This study intersects with several strands of economic history research concerning technological change, human capital, migration, and social mobility in Sweden.

**Electrification, Productivity, and Labor Markets in Sweden:**

The profound impact of electricity on Swedish industrialization is well-established. Schön (2000) documents the rapid uptake of electricity as a share of industrial energy use, particularly from 1900-1930, displacing traditional fuels and enabling new production methods [cite: 1 (Schon (2000) Electricity, technological change and productivity in Swedish industry, 1890-1990.pdf)]. He argues that electrification was driven both by falling relative prices and by technological shifts demanding electricity's specific qualities, noting its complementarity with skills and its role in structural transformation periods following crises [cite: 1 (Schon (2000) Electricity, technological change and productivity in Swedish industry, 1890-1990.pdf)]. Building on this macro-level understanding, Jayes et al. (2024) utilize the geographic specifics of the grid rollout to identify causal impacts at the individual level [cite: 1 (02-power-for-progress.pdf)]. Their difference-in-differences approach, comparing parishes along the "Western Line" to control parishes, confirms electricity's positive income effects and its inequality-reducing nature, benefiting less-educated workers and suggesting it complemented rather than replaced labor across various sectors [cite: 1 (02-power-for-progress.pdf)]. This paper takes the geographic setting from Jayes et al. (2024) but shifts the focus from the general labor force outcome to the characteristics of the high-skilled *inputs* – the engineers migrating into these specific locales.

**Human Capital, Skills, and Migration:**

The role of human capital, particularly engineering expertise, is crucial in adopting and leveraging new technologies like electricity [cite: 1 (Gronberg (2003) Learning and Returning Return Migration of Swedish Engineers from the United States, 1880–1940.pdf), 1 (Schon (2000) Electricity, technological change and productivity in Swedish industry, 1890-1990.pdf)]. While Jayes et al. (2024) showed benefits accruing to lower-skilled incumbents [cite: 1 (02-power-for-progress.pdf)], this paper examines the highly educated engineers essential for the system's expansion and operation. Their mobility aligns with findings on migration as a response to economic opportunity and a driver of skill acquisition. Andersson and Molinder (2024) demonstrate that during Sweden's industrialization (1880s-1930s), rural-urban migration significantly increased workers' skills (measured by occupational income), attributing roughly half the urban-rural skill gap to selection and half to the direct impact (both static matching and dynamic learning) of cities [cite: 1 (UPEH202412 (2).pdf)]. They find this effect was particularly strong for initially lower-skilled individuals, highlighting cities as engines of upward mobility [cite: 1 (UPEH202412 (2).pdf)]. Our study complements this by examining the characteristics of *high-skilled* migrants moving between specific types of *rural/semi-rural* parishes experiencing differential timing of a major technological shock, rather than the general rural-to-urban flow. We ask if the 'pull' of the earliest electrified areas selected engineers differently based on their pre-migration attributes.

**Elite Formation and Social Mobility:**

The background of economic elites is a central theme in understanding social structures and mobility. Henrekson, Lyssarides, and Ottosson (2021) investigate the social origins of CEOs of Sweden's 30 largest public firms since 1945 [cite: 1 (The social background of elite executives- the Swedish case Magnus Henrekson, Odd Lyssarides & Jan Ottosson (2021).pdf)]. They find a significant overrepresentation from the highest social group (Social Group I), although mobility increased somewhat for cohorts born in the mid-20th century, particularly from Social Group II [cite: 1 (The social background of elite executives- the Swedish case Magnus Henrekson, Odd Lyssarides & Jan Ottosson (2021).pdf)]. Their work confirms persistence at the very top, echoing findings that income mobility decreases at the highest percentiles [cite: 1 (The social background of elite executives- the Swedish case Magnus Henrekson, Odd Lyssarides & Jan Ottosson (2021).pdf)]. This paper contributes by examining a specific professional elite – engineers – during an earlier period of transformation. By comparing engineers moving to early versus later electrifying areas, we explore whether access to nascent technological opportunities was influenced by social origin, complementing studies like Henrekson et al. (2021) which look at established corporate leadership [cite: 1 (The social background of elite executives- the Swedish case Magnus Henrekson, Odd Lyssarides & Jan Ottosson (2021).pdf)]. Our finding of broad similarity across social origins for both groups suggests that, within this mobile technical elite, the *specific* location choice related to electrification timing was not strongly dictated by parental background.

**International Experience and Knowledge Transfer:**

The role of international experience, particularly from the United States and Germany, in shaping the skills and careers of Swedish engineers during this period is well-documented by Grönberg (2003) [cite: 1 (Gronberg (2003) Learning and Returning Return Migration of Swedish Engineers from the United States, 1880–1940.pdf)]. He characterizes many emigrating engineers as "target migrants," seeking specific knowledge and experience abroad with the intention of returning, leveraging this "symbolic capital" for career advancement in Sweden [cite: 1 (Gronberg (2003) Learning and Returning Return Migration of Swedish Engineers from the United States, 1880–1940.pdf)]. Grönberg notes the high return rates (over two-thirds overall, higher from Europe than the US) and the importance of engineers with foreign experience in key positions at major Swedish companies like ASEA and Sandviken, particularly in transferring knowledge about mass production, rational organization, and specific technologies [cite: 1 (Gronberg (2003) Learning and Returning Return Migration of Swedish Engineers from the United States, 1880–1940.pdf)]. He finds American experience was particularly valued, often compensating for lower educational levels or social origins, though the highest positions often went to those combining high status/education *with* foreign experience [cite: 1 (Gronberg (2003) Learning and Returning Return Migration of Swedish Engineers from the United States, 1880–1940.pdf)]. Our finding that engineers moving to the early-electrifying Western Line parishes were distinguished primarily by slightly more US experience aligns directly with Grönberg's emphasis on the perceived value of American know-how during this technological transition [cite: 1 (Gronberg (2003) Learning and Returning Return Migration of Swedish Engineers from the United States, 1880–1940.pdf)]. It suggests that while general engineering talent was mobile and responsive to broad regional dynamism, experience from the perceived technological frontier (the US) offered a marginal advantage or attraction specifically to the earliest adopting locales.

***

@@

In Sweden, Andersson and Molinder (@andersson2024did) investigated how migration influenced skill acquisition during industrialization (1880s–1930s). Using data from the Historical Swedish Population Panel (HISP) and linked censuses, they found that rural-urban migrants significantly increased their occupational income — a proxy for skills — after moving to cities. The impact of cities on skills had both static and dynamic effects. The static effect came from immediate access to higher-skilled jobs, while the dynamic effect reflected skill accumulation over time due to urban learning environments. Migrants with below-median initial incomes benefited the most, highlighting urban areas as drivers of upward mobility. Surprisingly, skill gains were similar across city sizes, including Stockholm and smaller urban centers. This suggests that the benefits of urbanization during Sweden's industrialization stemmed from the general characteristics of urban labor markets rather than city scale.

Heikkuri and Prado (@heikkuri2024technological) examined the impact of electrification on Swedish industry between 1913 and 1926. Using establishment-level data from industrial surveys, they found a positive correlation between electricity adoption, higher employment, and increased wages. Establishments that adopted electricity grew more rapidly in both employment and wages compared to those that did not, highlighting electrification as a key driver of economic advancement.  While direct evidence on skills is lacking, Heikkuri observed occupational shifts from farm laborers to factory workers in electrified parishes. Although both occupations were considered unskilled, factory workers earned significantly higher wages. This aligns with the findings in @jayes_power_2024 and suggests that electrification influenced labor reallocation and improved earnings for unskilled workers. However, Heikkuri and Prado also caution that factors such as collective bargaining agreements may have played a role, underscoring the need for further research to determine electrification's impact on skill demand.

@@

Some interesting papers that incorporate prediction and classification problems include @Bandiera2020ceo, who classify CEO behavior by collecting high-frequency diary information and then use a machine learning algorithm to classify CEOs into 'leaders' and 'managers' by the content of their meetings. @koschnick2023breaking uses a machine learning topic model to classify each paper by the universe of all students at English universities in the seventeenth and early eighteenth century to calculate a measure of how innovative the paper was; how it differed from the papers before it in the field and how similar the papers afterwards were. @OccC2024breaking in a paper titled "Breaking the HISCO Barrier: AI and Occupational Data Standardization" apply a neural network to the task of classifying an occupational description and benchmark their results against human labelling to show that the neural network achieved comparable accuracy with human labelling and involved an order of magnitude fewer human hours. I make use of this tool in the classification of occupations in this paper.

@@

## Materials and Methods

@@

### Source criticism

Obviously, the biographical dictionaries are not without their limitations. The criteria for inclusion in the dictionaries was somewhat vague, and individuals could opt in to being included for a nominal fee. It is not a representative sample of the Swedish population at the time. However, there should be some way to assess the representativeness of the biographical dictionaries. One way to do this is to compare the individuals in the biographical dictionaries to a different source of biographical information.

To assess the representativeness of the biographical dictionaries, I sampled a random selection of individuals from these dictionaries and manually checked whether they appear in the Swedish Biographical Lexicon (SBL).

The Swedish Biographical Lexicon (SBL), a personal history lexicon first published in 1917, now (as of spring 2024) includes entries from A to Södersten. It is a general inventory based on primary sources documenting important figures in Swedish society since about the 1500s, comprising approximately 27,200 pages with detailed data on individuals. The SBL aims to include individuals deemed significant by its editors, relying on broad and first-hand sources, without self-selection bias. This approach contrasts with the Vem är Vem? biographical dictionaries, where inclusion was voluntary, with criteria described as "somewhat vague," and individuals could pay to be featured. As a result, Vem är Vem? likely overrepresents those with a certain social standing and willingness to invest in being included, potentially limiting its representativeness of the wider society at the time.

I sampled 200 individuals from Vem är Vem? and found that only 38 (19%) were also present in the SBL, indicating that while Vem är Vem? contains a significant number of entries (about 75,000 individuals, representing roughly 1% of the Swedish population in the mid-20th century), it might overemphasize individuals of higher social status.

The level of detail provided in Vem är Vem? entries is inconsistent. Some individuals have extensive information on their careers, travels, publications, and organizational memberships, whereas others have only basic biographical information. This variability complicates the analysis of certain aspects of individuals' lives, particularly when attempting to classify career trajectories or sectors of work. However, while the amount of information might vary, we can clearly indicate when information is missing for individuals we analyze, and being transparent about these gaps is useful in understanding the limitations of our data.

@@

In order to analyze the biographical dictionaries, we need to bend the text into a machine readable structure. This process is not complicated, but somewhat involved. It includes breaking each component of the source up (e.g. each biography), extracting the pertinent information from each record, storing each value with its associated key, and then saving this information in a way that is easy to analyze and aggregate.

@@

I make use of the computational backend of Chat-GPT, a model called GPT-4o-mini from OpenAI to structure the information from the dictionaries and catalogue into a JSON format that I can analyze, step 3, as shown in @fig-data-collection-process. Many other LLMs, some open source, are available. I have chosen GPT-4o-mini as it is simple to interact with it in the programming language Python, and because doing so is relatively affordable for contained workloads such as this, when compared to hosting such a model on your own, beefy, costly computer.

To structure the 75,000 biographies in the dictionaries, it took about 15 hours of compute time, and cost 90 USD. When I timed myself extracting 10 biographies by hand, it took about 30 minutes. As such, the LLM method replaced 416 hours of human work with 15 hours of compute time, and 90 USD, plus the time invested in learning how to program the process.

@@

The GPT-4o-mini model which I make use of is a LLM which has been trained on all of Wikipedia and Wikidata, among other training material. These two sources contain the same information, but in a different format, as shown in the adapted extracts below; the text and @fig-wiki mimic the kind of material that the model I use is trained on.

As the base model underlying GPT-4o-mini is pre-trained to predict the next token on this kind of data, it has developed the ability through repeated exposure to this kind of biographic data to produce structured information from free text, and likewise construct natural sounding text from pieces of structured information. With the addition of *JSON mode* to the model, it is able to output structured information in a reliable manner as detailed in @openai2024text.

@@

Due to the large amount of material that GPT-4o-mini has been exposed to in training, it is familiar with the kinds of biographical text that I want it to structure. In order to draw on its familiarity with this kind of material, I need to provide it context about the biographical text it is being passed, and ensure that the model returns output in a useful way. I explain these prompts below.

@@

Next I explain the kinds of information that I want it to extract, and the exact format that I want it in. I do this by specifying what is called a Pydantic data model. This model provides fields that the API must return and specifications for the kind of data in each field. An except of the data model is shown in @fig-schema. The advantage of the Pydantic data model is that it allows me to specify the output in a way that is easy to read as a human, and a way in which OpenAI have built their API to accept. Since switching to this method of specifying the output, as opposed to a JSON schema, I have found that the percentage of successfully extracted biographies has increased from 80% to 94%.^[This is a rough estimate based on a sample of 1000 biographies that I manually scored.]

```{r}
#| label: fig-schema
#| fig-cap: "Excerpt of structuring schema"
#| fig-width: 5
#| out-width: 3in

knitr::include_graphics("assets/schema-3.png")
```

@@

`prompt = f"""`
`You are an expert on Swedish biographies and will structure the biographies of individuals from the 20th century biographical dictionary 'Vem är Vem' that is provided below.`

`### Task:`
`1. Use the schema to organize the information in the biography.`
`2. Keep the biographic descriptions in Swedish and remove any abbreviations based on your knowledge, e.g. 'fil. kand.' is 'filosofie kandidat', and 'Skarab. l.' is 'Skaraborgs Län' etc. `
`3. For missing data in a required field, include the field with a `None` value.`
`4. Ensure fields are correctly labeled and structured as per the schema.`
`5. Put years in full based on context. Put dates in DD-MM-YYYY format where possible. `

`"""`

@@

person_info <- tribble(
  ~key, ~value,
  "full_name", "Nordell, Axel Verner",
  "location", "Motala, Östergötland",
  "occupation", "Civilingenjör, kraftverksdirektör",
  "birth_date", "15/08/1881",
  "birth_place", "S. Möckleby, Kalmar",
  "birth_parents", "Gustaf N. and Amanda Seillergren",
  "parents_occupation", "Kyrkoherde",
  "birth_latitude", "56.35646300000001",
  "birth_longitude", "16.420155",
  "education_degree", "Studentexamen",
  "education_year", "1899",
  "education_institution", "Lunds högre allmänna läroverk",
  "education_latitude", "55.7046601",
  "education_longitude", "13.1910073"
)

@@

## Clustering and classification of occupations

The next task required classifying each individual in the biographical dictionaries into a particular occupation, and extracting the same information for a parent, if it is available.

For this task, I leaned on the tools of text embeddings, and a combination of unsupervised machine learning and advanced language processing techniques. I made use of a text embedding model trained on Swedish language text by the @bert-base-swedish-cased. It is an adaptation of the breakthrough BERT model, introduced by Google Research in 2018 [@google2016bert]. The advantage of this KB Lab model is that it has been trained on a selection of Swedish data, including books, news reports, and internet forums. Hence it is able to score the similarity of Swedish business descriptions and occupational titles.

@@

In @jayes_power_2024, we grouped occupations into three categories; direct electricity jobs (e.g. electricians), indirect electricity jobs that could benefit from electric motors (e.g. textile workers), and all other jobs. We made this classification based on the occupational title listed in the 1930 census alone. These titles are frequently used in economic history, along with a schema that classifies each title according to a list of possible titles and occupational descriptions. A widely used example is the Historical International Standard Classification of Occupations, or HISCO, defined originally by @vanLeeuwen2002HISCO. A wealth of mappings have been created that link an occupational string like *civilingenjör* to HISCO code 022, civil engineers, described as:

@@

While this classification process used to involve a large amount of human hours, several new methods that use machine learning have emerged to make this process more efficient. @OccC2024breaking use a neural network to classify occupational strings using a supervised machine learning process based on the occupational titles in several waves of Danish census data. @merouani2023innovation uses an unsupervised approach based on text embeddings to classify french occupational titles according to the HISCO schema, using a clever compartmentalization strategy to improve classification when individuals hold multiple titles.

In this paper I take a hybrid approach, making use of the additional data from the occupational trajectories of each individual; where they worked and what position they held at each firm or institution. This additional data helps me go beyond just a single occupational string and classify the sector that each engineer worked in. In many instances, the description of the individual in the biographical data relates to their education, rather than what they have done with it.

@@

## Descriptive statistics on structured data
